\funcitem \cppinline|string strn(T v)| \itt{strn}

\cppinline|string strn_sci(T v)| \itt{strn_sci}

These functions will convert the value \cppinline{v} into a string. This value can be of any type, as long as it is convertible to string. In particular, \cppinline{v} can be a vector, in which case the output string will contain all the values of the vector, separated by commas, and enclosed inside curly braces \cppinline{"{...}"}.

\begin{example}
\begin{cppcode}
strn(2);            // "2"
strn(true);         // "1"
strn("foo");        // "foo"
strn(vec1i{2,5,9}); // "{2, 5, 9}"
\end{cppcode}
\end{example}

The second version is dedicated to floating point numbers, and will output them in scientific format.

\begin{example}
\begin{cppcode}
strn_sci(2.0);    // "2.000000e+00"
strn_sci(2e10);   // "2.000000e+10"
strn_sci(-5e-2);  // "-5.000000e-02"

// Integer numbers are not affected
strn_sci(1000);   // "1000"
strn_sci(1000.0); // "1.000000e+03"
\end{cppcode}
\end{example}

\funcitem \cppinline|vec_<D,string> strna(vec<D,T>)| \itt{strna}

\cppinline|vec_<D,string> strna_sci(vec<D,T> v)| \itt{strna_sci}

These functions are the vectorized version of \cppinline{strn} (see above). The reason why the name of the function is different is because it is already possible to call \cppinline{strn} with vector arguments: the whole vector will be converted into a \emph{single} string. These versions however convert each element separately to form a vector of strings.

\begin{example}
\begin{cppcode}
strna(vec1i{1,5,6,9}); // {"1", "5", "6", "9"}
strn(vec1i{1,5,6,9});  // "{1, 5, 6, 9}"
\end{cppcode}
\end{example}

\funcitem \vectorfunc \cppinline|bool from_string(string s, T& v)| \itt{from_string}

This function tries to convert the string \cppinline{s} into a C++ value \cppinline{v} and returns \cpptrue in case of success. If the string cannot be converted into this value, for example if the string contains letters and the value has an arithmetic type, or if the number inside the string is too big to fit inside the C++ value, the function will return \cppfalse. In this case, the value of \cppinline{v} is undefined.

The vectorized version of this function will try to convert every value inside the string vector \cppinline{s}, and will store the converted values inside the vector \cppinline{v} (it will take care or properly resizing the vector, so you can pass an empty vector if you want). The return value will then be an array of boolean values, corresponding to the success or failure of each individual value inside \cppinline{s}. If an element of \cppinline{s} fails to convert, the corresponding value in \cppinline{v} will be undefined.

\begin{example}
\begin{cppcode}
float f;
bool b = from_string("3.1415", f);
b; // true
f; // 3.1415

b = from_string("abcdef", f);
b; // false;
f; // ??? could be 3.1415, or NaN, or anything else

vec1f fs;
vec1b bs = from_string({"1", "1.00e5", "abc", "1e128", "2.5"}, fs);
bs; // {true, true, false, false, true}
fs; // {1,    1e5,  ???,   ???,   2.5}
\end{cppcode}
\end{example}

\funcitem \cppinline|string hash(...)| \itt{hash}

This function scans all the arguments that are provided, and returns the hexadecimal representation of the SHA-1 hash of this argument list. The hash is a string such that: 1) all further calls of \cppinline{hash(...)} with arguments that have the exact same value will always return the same string, and 2) the probability that the function returns the same string for another set of arguments is very small (in fact, although this algorithm was created in 1995, no such collision has been found yet).

This is useful for example to cache the result of some heavy computation: once the computation is done, the input parameters of the computation can be fed to \cppinline{hash()} to give a ``sort-of-unique'' identifier to the result, and this result can be saved somewhere with this identifier. Later on, if the computation is requested with another set of parameters, they are fed to \cppinline{hash()} and the resulting string is compared to all the saved results: if a match is found, then the associated pre-computed result can be used immediately, else the computation goes on.

\begin{example}
\begin{cppcode}
std::string s;

// With a single argument
s = hash("hello world!");
s; // "da52a1357f3c973e1ffc1b694d5308d0abcd9845"
s = hash("hello world?")
s; // "793e673d04e555f8f0b38033d5223c525a040719"
// Notice how changing a single character gives a completely
// different hash string

// With multiple arguments
s = hash(1, 2, 3);
s; // "570331ab965721aae8a8b3c628cae57a21a37560"
s = hash("123");
s; // "0e898437b29ec20c39ca48243e676bcb177d4632"
s = hash(1.0, 2.0, 3.0);
s; // "9c45014f7c7943cb7860f3db4b885fb44b510ec8"
\end{cppcode}
\end{example}
